Music Recommendation System Using the Million Song Dataset

### 2015-05-08 writeup/MSDgroupwriteup.tex
Update MSDgroupwriteup.tex
Updating .tex

After testing various values of the parameters, it was soon discovered that matrix factorization results were significantly below the popularity baseline and changing the hyper-parameters around did not help much. The best MAP value of 0.0038 was achieved using a high rank $d$ of 80 over 100 iterations with a variance of 1.0. The log joint likelihood chart can be used to verify that the probabilistic matrix factorization algorithm is running properly in that the values are monotonically decreasing which corresponds to a decreasing training error. The plot shows that there is more room for improvement of the training error, but the MAP plot that was taken from the same run shows that there is only a marginal difference in terms of improving the MAP score after the initial 5 iterations. Due to this result, further exploration of the probabilistic matrix factorization algorithm was limited to 30 iterations to reduce computational time.

@@

One explanation for the very low MAP scores using PMF was that the number of plays that were used are different from what PMF is typically used for which is bounded ratings. Using implicit user feedback (number of plays) instead of explicit user feedback (ratings) has been found in literature to create various issues. One of the main issues is that implicit feedback is inherently noisy and does not measure negative feedback. Users may have played a song to simply find out that they dislike the song. The other main issue is that the exact number of plays could have a significantly different meaning across different users. For example, an avid user who has played a song five times could have a different meaning as compared to a user that has at most only played songs five times.

@@

From the experiments, it was found that the rank $d$ did not contribute significantly to the MAP values especially in settings where the MAP values were close to the baseline. Higher variance values were found to actually significantly increase the MAP values. In terms of scaling schemas, making the data binary produced the best result, with the second best being normalizing the data against the total number of plays for each user. Perhaps it can be said that using number of plays introduces more noise than simply considering if the user has played the song or not. Interestingly enough, normalizing the data against total number of plays for each song performed the worst and this most likely relates to the fact that number of plays is more meaningful relative to each user rather than to each song.

@@

Regardless, the best result achieved for PMF in the end was 0.0143, which was only slightly above the popularity baseline. Based on literature review, we have found that this is better or in line with what others have achieved with using PMF on play counts. Nevertheless, this indicates that PMF is not suitable for the MSD dataset and this can be attributed to the implicit feedback issue or simply to the fact that the dataset is too sparse. However, it is still possible to apply matrix factorization techniques to this dataset through using metadata and this has been demonstrated to be viable by \citet{liangcodebook} which achieved MAP values exceeding 0.1.

\citep{hu2008collaborative}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{lc}
\hline
\bf{Algorithm} &  \bf{MAP score}\\ \hline
Item Based  CF  &  0.0479 \\ 
Artist Based Popularity Baseline  & 0.0448    \\ 
User Based  CF  &  0.0377 \\ 
Matrix Factorization  &   0.0143  \\ 
Top 500 Songs by Count  Baseline &  0.0138  \\ 
Top 500 Songs by Plays  Baseline &  0.0126  \\ 
\end{tabular}
\end{center}
\end{table}

>>>>

After testing various values of the parameters, it was soon discovered that matrix factorization results were significantly below the popularity baseline and changing the hyper-parameters around did not help much. The best MAP value of 0.0039 was achieved using a high rank $d$ of 80 over 100 iterations with a variance of 1.0. The log joint likelihood chart can be used to verify that the probabilistic matrix factorization algorithm is running properly in that the values are monotonically increasing, which corresponds to a decreasing training error. The first graph in Figure /ref{defaultPMF} shows that the log joint likelihood is indeed monotonically increasing and slowed down significantly after the first twenty iterations. The second graph in the same figure shows the MAP values from the same run and it shows that the MAP values decreasing after the maximum at five iterations. It could be argued that emphasis should be placed on finding the right number of iterations to produce the best MAP values, but since the overall MAP results were poor and the differences were minimal, it was decided that this was not to be a priority for further experimentation. Due to this conclusion, further exploration of the probabilistic matrix factorization algorithm was limited to 30 iterations in order to reduce computational time.

@@

One explanation for the very low MAP scores using PMF was that the number of plays that were used are different from what PMF is typically used for which is bounded ratings. Using implicit user feedback (number of plays) instead of explicit user feedback (ratings) has been found in literature \citep{hu2008collaborative} to create various issues. One of the main issues is that implicit feedback is inherently noisy and does not measure negative feedback. Users may have played a song to simply find out that they dislike the song. The other main issue is that the exact number of plays could have a significantly different meaning across different users. For example, an avid user who has played a song five times could have a different meaning as compared to a user that has at most only played songs five times.

@@

The results from the experiments were plotted in Figure \ref{PMF}. It was found that the rank $d$ did not contribute significantly to the MAP values especially in settings where the MAP values were close to the popularity baseline. Higher variance values were found to actually significantly increase the MAP values and it can be argued that the reason why the default PMF works best at variance of 10 instead of 1.0 for the others is that the empirical variance of play count values is at least an order of magnitude higher than the others. This also suggests that the variance used in PMF should scale with the empirical variance, but the exact relationship remains unclear.

In terms of scaling schemas, making the data binary produced the best result, with the second best being normalizing the data against the total number of plays for each user. Perhaps it can be said that using number of plays introduces more noise than simply considering if the user has played the song or not. TF-IDF performed slightly worse than the two, so down-weighting the songs shared more often by users did not produce better recommendations. Interestingly enough, normalizing the data against total number of plays for each song performed the worst and this most likely relates to the fact that number of plays is more meaningful relative to each user rather than to each song.

@@

Regardless, the best result achieved for PMF in the end was 0.0143, which was only slightly above the popularity baseline. Based on literature review \citep{li20122million, McFee:2012:MSD:2187980.2188222}, we have found that this is better or in line with what others have achieved with using PMF on play counts. Nevertheless, this indicates that PMF is not suitable for the MSD dataset and this can be attributed to the implicit feedback issue or simply to the fact that the dataset is too sparse. However, it is still possible to apply matrix factorization techniques to this dataset through using metadata and this has been demonstrated to be viable by \citet{liangcodebook} which achieved MAP values exceeding 0.1.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{lc}
\hline
\bf{Algorithm} &  \bf{MAP score}\\ \hline
Item Based  CF  &  0.0479 \\ 
Artist Based Popularity Baseline  & 0.0448    \\ 
User Based  CF  &  0.0377 \\ 
Matrix Factorization  &   0.0143  \\ 
Top 500 Songs by Count  Baseline &  0.0138  \\ 
Top 500 Songs by Plays  Baseline &  0.0126  \\ 
\end{tabular}
\caption{Final MAP results across various algorithms.}\label{tab:stats}
\end{center}
\end{table}
